Political Polarization, Public Trust, and Healthcare Disparities: Unraveling COVID-19's Impact and Future Preparedness

### 2024-02-15 outputs/paper/paper.qmd
Final Edits

Our analysis is grounded in comprehensive data sourced from the paper, **"Why Did the Best Prepared Country in the World Fare So Poorly during COVID?"** [@citearticle], enabling a nuanced exploration of how community health systems and socio-political contexts intersect to influence disease outcomes. While leveraging the initial findings from @citearticle, we expand the scope of inquiry to investigate broader implications and unearth deeper insights into the complexities of COVID-19 disparities. We particularly focus on the impact of political polarization on adherence to health guidelines and the role of social vulnerabilities such as healthcare access and housing conditions on the life expectancy of various racial and ethnic groups. One key finding that emerges from our study is the influence of political polarization, with Republican-leaning states exhibiting slightly higher clustering around elevated death rates, highlighting the importance of socio-political factors in shaping public health responses.

@@

As we navigate the complexities of the COVID-19 pandemic, it is imperative to not only understand the factors contributing to disparities but also to identify actionable strategies for mitigating these inequities and strengthening community resilience. In the subsequent sections of this paper, we outline our research process, findings, and implications as follows: @sec-data delves into our methodology, encompassing data collection, cleaning, and measurement techniques. In @sec-results, we present detailed findings accompanied by graphs, tables, and other visual aids for clarity. Following this, @sec-discussion engages in a discussion on the significance and implications of our results for public health policy and practice, addressing unexpected findings, limitations, and avenues for further research. Finally, @sec-conclusion offers a conclusion and recommendations, summarizing key insights and providing actionable recommendations for policymakers, practitioners, and future research directions. By elucidating these critical relationships and expanding upon prior research, our study contributes to the broader discourse on pandemic preparedness and underscores the importance of tailored interventions in addressing underlying structural determinants of health disparities.

@@

Incorporating this understanding into our analysis, our measurement for COVID death rates is drawn from an accumulated dataset last updated until March 2023. This temporal constraint means we would not be aware of the fluctuations of death rates over time, particularly during the peak of the pandemic in 2020. The data's primarily reliance on the reported statistics from hospitals, clinics, testing centres, and laboratories may introduce bias as excludes uncounted deaths from COVID. Especially during the critical period of health facility capacity constraints, many individuals died outside of medical facilities without being documented. Therefore, our measurements for COVID death rates only accounted for the documented deaths from COVID.

@@

Manual Data Extraction involved careful review of reports and tables to ensure accurate data capture, particularly for datasets like life expectancy where this meticulous approach was paramount; this was coupled with Standardization and Adjustment processes where excess death data and age-standardized excess death rates underwent significant processing, including the comparison of observed deaths during the pandemic to expected deaths based on historical trends, with adjustments made for the age distribution of the population; concurrently, Geospatial Data Handling was employed for mapping the GHS Index scores, utilizing Geographic Information Systems (GIS) to visually represent pandemic preparedness across various countries, and this was complemented by Real-Time Data Aggregation, which entailed aggregating COVID-19 death data in real-time from authoritative sources such as the Johns Hopkins University Dashboard, necessitating continuous data monitoring and updates to maintain accuracy and relevance.

@@

Given the variation in population sizes among states, direct comparisons of COVID-19 case numbers are inherently flawed. As an alternative, we employed death rates per 100,000 people as a metric for evaluating each state's COVID preparedness and situation. We then produced @tbl-deaths that rank the top ten states with the highest death rates from COVID-19 per 100,000 people to examine the potential correlation between party preferences and COVID-related deaths. Notably, We found that six of the ten top Republican states made a reappearance in the top death rates table; these states are Oklahoma, West Virginia, Arkansas, Alabama, Tennessee, and Kentucky. The higher mortality rates observed in Republican-leaning states may be attributed to their preference for less stringent measures compared to Democratic-leaning states, which tend to favor stricter measures [@vandusky-allen_shvetsova_2021].

>>>>

Grounded in comprehensive data sourced from the paper, **"Why Did the Best Prepared Country in the World Fare So Poorly during COVID?"** [@citearticle], our analysis offers a nuanced exploration of how community health systems and socio-political contexts intersect to influence disease outcomes. While leveraging the initial findings from @citearticle, we expand the scope of inquiry to investigate broader implications and unearth deeper insights into the complexities of COVID-19 disparities. We particularly focus on the impact of political polarization on adherence to health guidelines and the role of social vulnerabilities such as healthcare access and housing conditions on the life expectancy of various racial and ethnic groups. One key finding that emerges from our study is the influence of political polarization, with Republican-leaning states exhibiting slightly higher clustering around elevated death rates, highlighting the importance of socio-political factors in shaping public health responses.

@@

As we navigate the complexities of the COVID-19 pandemic, it is imperative to not only understand the factors contributing to disparities but also to identify actionable strategies for mitigating these inequities and strengthening community resilience. In the subsequent sections of this paper, we outline our research process, findings, and implications as follows: @sec-data delves into our methodology, encompassing data collection, cleaning, and measurement techniques. In @sec-results, we present detailed findings accompanied by graphs, tables, and other visual aids for clarity. Following this, @sec-discussion engages in a discussion on the significance and implications of our results for public health policy and practice, addressing unexpected findings, limitations, and avenues for further research. Finally, @sec-conclusion offers a conclusion that synthesizes key insights from our analysis. By elucidating these critical relationships and expanding upon prior research, our study contributes to the broader discourse on pandemic preparedness and underscores the importance of tailored interventions in addressing underlying structural determinants of health disparities.

@@

Incorporating this understanding into our analysis, our measurement for COVID death rates is drawn from an accumulated dataset last updated until March 2023. This temporal constraint means we would not be aware of the fluctuations of death rates over time, particularly during the peak of the pandemic in 2020. The data's primary reliance on the reported statistics from hospitals, clinics, testing centres, and laboratories may introduce bias as it excludes uncounted deaths from COVID. Especially during the critical period of health facility capacity constraints, many individuals died outside of medical facilities without being documented. Therefore, our measurements for COVID death rates only accounted for the documented deaths from COVID.

@@

Manual Data Extraction involved careful review of reports and tables to ensure accurate data capture, particularly for datasets like life expectancy where this meticulous approach was paramount. This was coupled with Standardization and Adjustment processes where excess death data and age-standardized excess death rates underwent significant processing, including the comparison of observed deaths during the pandemic to expected deaths based on historical trends, with adjustments made for the age distribution of the population. Concurrently, Geospatial Data Handling was employed for mapping the GHS Index scores, utilizing Geographic Information Systems (GIS) to visually represent pandemic preparedness across various countries, and this was complemented by Real-Time Data Aggregation, which entailed aggregating COVID-19 death data in real-time from authoritative sources such as the Johns Hopkins University Dashboard, necessitating continuous data monitoring and updates to maintain accuracy and relevance.

@@

Given the variation in population sizes among states, direct comparisons of COVID-19 case numbers are inherently flawed. As an alternative, we employed death rates per 100,000 people as a metric for evaluating each state's COVID preparedness and situation. We then produced @tbl-deaths, which ranks the top ten states with the highest death rates from COVID-19 per 100,000 people, to examine the potential correlation between party preferences and COVID-related deaths. Notably, we found that six of the ten top Republican states made a reappearance in the top death rates table; these states are Oklahoma, West Virginia, Arkansas, Alabama, Tennessee, and Kentucky. The higher mortality rates observed in Republican-leaning states may be attributed to their preference for less stringent measures compared to Democratic-leaning states, which tend to favor stricter measures [@vandusky-allen_shvetsova_2021].
